---
title: "Homework II"
author: "J Lacasa"
date: "2026-02-15"
output:
  pdf_document:
    number_sections: true
---

Read the exercises below, answer them, and knit the .Rmd file into an HTML of PDF file. Rename it, including your last name (e.g., "Smith_hw2.Rmd") and submit it on CANVAS by Tue, February 24th. 

The exercises below aim to (i) review the most important aspects/benefits of multilevel models, (ii) review the analysis of classic designed experiments, and (iii) review the analysis of any data with a clear/"obvious" data architecture. 

# Multi-level models

Consider a data set collected to study the water quality for different agricultural practices (e.g., conservation practices, intensive practices, etc.) across an area covering different watersheds. 
In each ($j$th) watershed, water samples were taken to measure nitrate concentrations for a field under the $i$th agricultural  practice. Note that multiple fields (with different agricultural practices) may fit in the same watershed. 
Then, we have multiple observations $y_{ijk}$ for the $i$th agricultural practice in the $j$th watershed and $k$th field. 
The main objective of the study is to quantify the water quality for the different agricultural practices. All watersheds are equally important and their size can be assumed similar, but some watersheds have more observations than others, and they also have different proportions of the presence of the agricultural practices.    

## Data architecture 

Draw a schematic representation of how the architecture in the data looks like and embed a picture of that representation in this document. 

## Recovery of inter-group information 

Consider the model 

$$y_{ijk} = \mu + P_i +w_j+ \varepsilon_{ijk}, \\ \varepsilon_{ijk} \sim N(0, \sigma^2),$$
where:

- $y_{ijk}$ is the observation of the $i$th agricultural practice in the $j$th watershed, and $k$th field, 
- $P_i$ is the effect of the $i$th agricultural practice, 
- $w_j$ is the effect of the $j$th watershed, and 
- $\varepsilon_{ijk}$ is the residual. 

Something we learned in this course is that the assumption behind $w_j$ can affect our results and our inference. In this case, how does modeling $w_j$ as a fixed effect versus as a random effect affect inference and results?



# Analysis of an incomplete block design 

The data below were generated by a designed experiment with a one-way treatment structure, where genotype is the treatment factor and has 13 levels, in an incomplete block design. 

```{r}
url <- "https://raw.githubusercontent.com/stat799/spring2026/refs/heads/main/data/bibd.csv"
dat <- read.csv(url)
head(dat)
```

## Model fitting 

Write a statistical model to describe the data generating process and fit said model using statistical software. 

Justify the different components of the model (e.g., fixed and/or random effects).

Make sure (and show) that the model is reliable, and show why it is reliable.


# Analysis of a split-plot designed experiment. 

The data below were generated for a study aiming to evaluate the individual plant performance (in grams of grain per plant) for different fertilizer sources (8 different sources) and moments of application (4 different sources). 
Scientists ran a designed experiment with a 8 $\times$ 4 factorial treatment structure, in a split plot in an RCBD. Fertilizer source was the whole-plot treatment factor, and the application moment was the split-plot treatment factor. 
Note that two plants were observed per plot (indicated as "sample"). 

```{r}
url <- "https://raw.githubusercontent.com/stat799/spring2026/refs/heads/main/data/splitplot_subs.csv"
dat <- read.csv(url)
head(dat)
```

## ANOVA 

Write the ANOVA table (or do the multilevel ANOVA from Gelman (2005) from day 4) including source of variability and degrees of freedom.  

## Marginal means 

Get the marginal means, and explain what sources of uncertainty are contained in the standard errors of the means. 

## Drawing conclusions 

The scientists that ran this experiment were mostly interested in finding out the best Fertilizer-moment combination that maximizes yield, while understanding what makes that combination the best (i.e., is the effect of fertilizer source or application moment more important? Are they equally important?)




